---
title: "Retrotransposons analysis"
author: "Christelle Colin-Leitzinger"
date: '`r Sys.Date()`'
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: false
    theme: united
    highlight: pygments
    df_print: paged
editor_options: 
  chunk_output_type: console
---

<style type="text/css">

.figure {
   margin-top: 25px;
   margin-bottom: 100px;
}

table {
    margin-top: 10px;
    margin-bottom: 25px;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      cache = FALSE,
                      fig.align='center'
                      )
```

```{r library, echo = FALSE}
library(tidyverse)
# library(REMP)
# library(ComplexHeatmap)
library(gtsummary)
library(survival)
library(survminer)
theme_gtsummary_compact()
theme_set(theme_classic())
```

<style>
div.darkblue { background-color:#0099CC; border-radius: 5px; padding: 20px; font-size: 38px}
</style>
<div class = "blue">

<span style="color: white;">Retrotransposons analysis in AACES and NCOCS data</span>

</div>
<br>

```{r load}
load(paste0(here::here(), "/cleaned_07082022.rda"))

cluster_res_list <- read_rds(paste0(here::here(), "/clustering/MOVICS/movics_cluster_res_list_08032023.rds"))

L1_imputed <- read_rds(paste0(here::here(), "/L1_imputed.rds"))
Alu_imputed <- read_rds(paste0(here::here(), "/Alu_imputed.rds"))
ERV_imputed <- read_rds(paste0(here::here(), "/ERV_imputed.rds"))
```


```{r cleaning}
coca_cluster_results <- cluster_res_list$COCA$clust.res %>%
  remove_rownames() %>% 
  mutate(patient_id = str_remove(samID, "X")) %>% 
  select(-samID, coca_RE_cluster = clust)

phenoclean <- phenoclean %>%
  select(-c(starts_with("chr"))) %>%
  remove_rownames() %>%
  mutate(suid = Sample_Name) %>%
  select(patient_id = Complete.Barcode, suid, everything(), -c(Sample_Name, roi_num : Basename))

phenoclean <- phenoclean %>%
  mutate(bmi_cat = case_when(
    BMI_recent < 25                                         ~ "<25",
    BMI_recent >= 25 &
      BMI_recent < 30                                       ~ "25-29",
    BMI_recent >= 30 &
      BMI_recent < 35                                       ~ "30-34",
    BMI_recent >= 35                                        ~ "≥35"
  ), bmi_cat =
    factor(bmi_cat, levels = c("<25", "25-29", "30-34", "≥35"))) %>%
  mutate(bmi_cat2 = case_when(
    BMI_recent < 30                                         ~ "<30",
    BMI_recent >= 30                                        ~ "≥30"
  ), bmi_cat2 =
    factor(bmi_cat2, levels = c("<30", "≥30"))) %>%
  mutate(os_event = case_when(
    vitalstatus == 1                ~ 0,
    vitalstatus == 2                ~ 1
  )) %>% 
  rename(os_time = timelastfu) %>% 
  mutate(smoking_status = case_when(
    str_detect(smokcurrent, "current|former") ~ "Ever",
    smokcurrent == "never smoker"             ~ "Never"
  ))

```


```{r prep data}
met_data <- phenoclean %>% 
  inner_join(., coca_cluster_results,
            by = "patient_id") %>% 
  mutate(across(where(is.character), ~ str_to_sentence(.)))
```



# I. Table 1. Patient characteristics overall
```{r}
met_data %>% 
  select(refage, race, ethnicity_dnam,
         karyotype,
         stage, histology, grade, 
         BMI_recent, bmi_cat, bmi_cat2,
         smokcurrent, smoking_status,
         debulking_CA125, neoadjuvant, adjuvant,
         vitalstatus2,
         coca_RE_cluster
  ) %>% 
  tbl_summary(
    statistic = list(all_continuous() ~ "{mean} ({sd})"),
    label = list(refage ~ "Age at diagnosis",
                 debulking_CA125 ~ "Debulking Status", 
                 neoadjuvant ~ "Neoadjuvant chemotherapy",
                 adjuvant ~ "Adjuvant chemotherapy",
                 smokcurrent ~ "Current smoking status"
    ),
    type = list(c(neoadjuvant, adjuvant) ~ "categorical"),
    digits = list(all_continuous() ~ 1)) %>% 
  add_stat_label() %>%
  modify_header(list(label ~ "**Patient characteristics**", 
                     all_stat_cols() ~ "**{level}**, N = {n}"))
```
<br>
<br>

# II. Table 2. Patient characteristics by cluster
```{r}
met_data %>% 
  select(refage, race, ethnicity_dnam,
         karyotype,
         stage, histology, grade, 
         BMI_recent, bmi_cat, bmi_cat2,
         smokcurrent, smoking_status,
         debulking_CA125, neoadjuvant, adjuvant,
         vitalstatus2,
         coca_RE_cluster
  ) %>% 
  tbl_summary(by = coca_RE_cluster,
              statistic = list(all_continuous() ~ "{mean} ({sd})"),
              label = list(refage ~ "Age at diagnosis",
                           debulking_CA125 ~ "Debulking Status", 
                           neoadjuvant ~ "Neoadjuvant chemotherapy",
                           adjuvant ~ "Adjuvant chemotherapy",
                           smokcurrent ~ "Current smoking status"
              ),
              type = list(c(neoadjuvant, adjuvant) ~ "categorical"),
              digits = list(all_continuous() ~ 1)) %>%
  bold_labels() %>% add_stat_label() %>% 
  add_p() %>% bold_p(t=.05) %>% add_overall() %>% 
  modify_spanning_header(c(stat_1, stat_2) ~ "COCA cluster") %>% 
  modify_header(list(label ~ "**Patient characteristics**"))
```

# III. Adjusted hazard ratios and 95% confidence intervals

## 1. Adjusted hazard ratios and 95% confidence intervals
```{r}
tbl1 <- met_data %>%
  select(coca_RE_cluster,
         refage, stage) %>%
  tbl_uvregression(method = survival::coxph,
                   y = (Surv(time = met_data$os_time,
                             event = met_data$os_event)),
                   exponentiate = TRUE) %>%
  bold_p(t = .05) %>% 
  add_nevent(location = "level") %>% add_n(location = "level") %>%
  bold_labels() %>% italicize_levels()
tbl2 <- coxph(Surv(time = met_data$os_time, 
                   event = met_data$os_event) ~ coca_RE_cluster + refage + stage, 
              data = met_data)  %>% 
  tbl_regression(exponentiate = TRUE) %>% 
  bold_p(t = .05) %>% 
  add_nevent(location = "level") %>% add_n(location = "level")

tbl_merge(list(tbl1, tbl2), tab_spanner = c("**Univariable**", "**Multivariable**"))
```
<!-- PH assumption -->
<!-- ```{r} -->
<!-- coxph(Surv(time = met_data$os_time, -->
<!--                    event = met_data$os_event) ~ coca_RE_cluster + refage + stage, -->
<!--       data = met_data)  %>% -->
<!--   cox.zph() -->
<!-- ``` -->
<br>
<br>


## 2. HR adjusted by debulking status
```{r}
tbl1 <- met_data %>%
  select(coca_RE_cluster,
         refage, stage, debulking_CA125) %>%
  tbl_uvregression(method = survival::coxph,
                   y = (Surv(time = met_data$os_time,
                             event = met_data$os_event)),
                   exponentiate = TRUE) %>%
  bold_p(t = .05) %>% 
  add_nevent(location = "level") %>% add_n(location = "level") %>%
  bold_labels() %>% italicize_levels()
tbl2 <- coxph(Surv(time = met_data$os_time, 
                   event = met_data$os_event) ~ coca_RE_cluster + refage + stage + debulking_CA125, 
              data = met_data)  %>% 
  tbl_regression(exponentiate = TRUE) %>% 
  bold_p(t = .05) %>% 
  add_nevent(location = "level") %>% add_n(location = "level")

tbl_merge(list(tbl1, tbl2), tab_spanner = c("**Univariable**", "**Multivariable**"))
```
PH assumption
```{r}
coxph(Surv(time = met_data$os_time, 
                   event = met_data$os_event) ~ coca_RE_cluster + refage + stage + debulking_CA125, 
      data = met_data)  %>% 
  cox.zph()
```
<br>
<br>

# IV. KM
```{r Survival, fig.height = 7}
ggsurvplot(survfit(Surv(os_time, os_event) ~ coca_RE_cluster, 
                   data=met_data),
           title = "OS Analysis",
           font.main = c(20, "bold", "black"),
           font.x = c(18, "bold", "black"),
           font.y = c(18, "bold", "black"),
           font.legend = c(16, "black"),
           font.tickslab = c(16, "bold", "black"),
           size = 1.5,

           xlab = "Time in months",
           legend = "top",
           legend.title = "",
           pval = TRUE,
           conf.int = FALSE,
           # Censor
           censor = TRUE
) + guides(colour = guide_legend(ncol = 1))
```
<br>
<br>

# V. Sensitivity analysis

## 1. Sensitivity analysis in patients who received adjuvant
```{r}
adj_data <- met_data %>%
  filter(adjuvant == "Yes")

coxph(Surv(time = adj_data$os_time,
                   event = adj_data$os_event) ~ coca_RE_cluster + refage + stage + debulking_CA125,
              data = adj_data)  %>%
  tbl_regression(exponentiate = TRUE) %>%
  bold_p(t = .05) %>%
  add_nevent(location = "level") %>% add_n(location = "level")
```
<br>
<br>

# 2. Sensitivity analysis in patients with serous histology
```{r}
serous_data <- met_data %>%
  filter(histology == "Serous")

coxph(Surv(time = serous_data$os_time,
                   event = serous_data$os_event) ~ coca_RE_cluster + refage + stage + debulking_CA125,
              data = serous_data)  %>%
  tbl_regression(exponentiate = TRUE) %>%
  bold_p(t = .05) %>%
  add_nevent(location = "level") %>% add_n(location = "level")
```

# 3. Sensitivity analysis in patients with Black or african american DNA
```{r}
black_data <- met_data %>%
  filter(ethnicity_dnam == "Black or african american")

coxph(Surv(time = black_data$os_time,
                   event = black_data$os_event) ~ coca_RE_cluster + refage + stage + debulking_CA125,
              data = black_data)  %>%
  tbl_regression(exponentiate = TRUE) %>%
  bold_p(t = .05) %>%
  add_nevent(location = "level") %>% add_n(location = "level")
```





